---
title: "Data preparation with MADRaT"
author: "Jan Philipp Dietrich"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data preparation with MADRaT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

MADRaT is a framework that can help structuring data preparation in R. It splits the data preparation into separate steps with each having distinctive requirements about the returned data. The following tutorial will describe the first steps with the package together with the specific requirements for each calculation step. 


# Setup

MADRaT requires a local directory to store data such as downloaded source data, cache files and output. Running `getConfig` in the package for the first time it will ask you for a folder to use and store that setting permanently (if allowed by the user).

```{r, echo = TRUE}
library(madrat)
cfg <- getConfig()
```

After setting that directory that package is ready to use. If not stated otherwise in the config, all downloaded source data and created output files can be found in the subdirectories "sources" or "output" of the main directory, respectectively.

# MADRaT framework components

MADRaT splits the process of data preparation into the following components. Each component comes with a MADRaT wrapper function managing the data preparation process and performing some sanity checks on the calculations. The wrapper functions will run user defined functions which are specific to a certain source or calculation.

## downloadSource

The first step in data preparation is downloading the source data. `downloadSource` will create a folder for the given source and set all local file paths correctly. The user defined download function must contain the code required to download the source data in to the local folder the script is run from. An example for such a function is `madrat:::downloadTau`.

```{r, echo = TRUE}
madrat:::downloadTau
```

The name of the user function always has to be a combination of the function type (in this case "download") and the source or calculation type (in this case "Tau"). The wrapper function always expects the source or calculation type as argument. To run downloadTau through the wrapper one has to use the following call:

```{r, echo = TRUE, eval=FALSE}
downloadSource("Tau", overwrite = TRUE)
```

Here we set `overwrite=TRUE` to make sure that the data is downloaded in any case. In the default case `overwrite=FALSE` data will only be downloaded if there is not already an existing source folder containing the data.

## readSource

As soon as the data is available in the source folder it can be read it. Reading is performed by `readSource` and is split into 1  to 3 steps (depending on the data): read, correct and convert.

### read

In the first step the data is read into R and converted to a magclass object. Except of the conversion no other modifications are performed and the content of the magclass object should be completely identical to the downloaded data.

```{r, echo = TRUE}
madrat:::readTau
```

Only reading in data can be done by running `readSource` with the argument `convert=FALSE`:

```{r, echo = TRUE, eval=FALSE}
x <- readSource("Tau", "paper", convert=FALSE)
```

If a data source comes with several files it is sometimes necessary to specify a subtype. In the given example the source data comes with two data sets ("paper" and "historical"). In the example above the subtype "paper" is chosen.

### correct

The correction step is optional and can be used to fix issues in the data such as removing duplicates, replacing NAs or other corrections. This step is purely about fixing quality problems in the input data. If this step is required one can create a correct-function such as `correctTau` for the data source "Tau". As the example data "Tau" does not require any of these corrections there is no correct function in the given example data.

### convert

To allow for flexible aggregation of data to world regions and for compatibility between different data sources MADRaT imposes a standard spatial resolution on all data sources. The used standard is the ISO 3166-1 3 digits country code standard. The function `getISOlist()` returns a vector of these countries. 

After conversion the data set should provide numbers for all countries listed in that standard. Te wrapper function `readSource` will throw an error if countries are missing. It is important that a best guess is used for countries which are not directly provided by the given source as everything else might lead to errors or critical biases in the follow up calculations. Support tools such as `toolCountryFill` help to interpolate the missing information:

```{r, echo = TRUE}
madrat:::convertTau
```

Read and convert can be run together by running `readSource`:

```{r, echo = TRUE, eval=FALSE}
x <- readSource("Tau", "paper")
```

As `correct` also the `convert` function is optional, but not providing is indicates to MADRaT that the resulting data is not on ISO country level and will therefor not be available for aggregation to world regions. Cases in which sources will not have a convert function are data sets without spatial resolution (e.g. providing only a global value) or data sets which should for other reasons not be aggregated to country level. For most cases a `convert` function should exist.

As the corrections performed in a `correct` function are usually similar to the interpolations performed in a `convert` function it is also possible to have these corrections just included in the `convert` functions. For this reason most sources usually have a `read` and a `convert` but not a `correct` function.

## calcOutput

Besides reading in a data source and preparing it for further usage data preparation often requires to extract certain information out of the given data sources. In contrast to the steps before this can also mean blending two or more data sets into one output data set. For this reason MADRaT distinguishes between the source type, which is always linked to a specific source, and a calculation type, which is always linked to a specific data output. 

In the given example the data source "Tau" is used to calculate a data output called "TauTotal".

```{r, echo = TRUE}
madrat:::calcTauTotal
```

calc-Functions always have to return a list of objects with some list entries mandatory and others optional. Mandatory entries are the calculated data object in magclass format `x` a `weight` for aggregating the data from country level to world regions (can be `NULL` if the data just should be summed up), a  short `description` of the data set and the `unit` of the data set. 
Optional statements are for instance a `note` with additional details about the data or `min` and `max` values for the data which will be used for sanity checking the data coming out of the calculation. A full overview about required and/or allowed list entries can be found in the help to `calcOutput` (`?calcOutput`).

An output calculation can be run with the wrapper function `calcOutput`:

```{r, echo = TRUE, eval=FALSE}
x <- calcOutput("TauTotal")
```

By default it will return the data aggregated to the world regions set in the MADRaT configuration. Adding the argument `aggregate=FALSE` will return the data in its original resolution and is useful if a calc function is used as source for another calc function.

## retrieveData

When preparing data for a certain purpose it is often the case that not only one but several data sets have to be prepared as a collection of data. This is where `retrieveData` steps in. It allows to create a collection of data sets and manages it calculation and packaging. The user defined functions matching to the wrapper `retrieveData` start with `full` in the name:

```{r, echo = TRUE}
madrat:::fullEXAMPLE
```

Each function must have the argument `rev` which contains a revision number. This can be used to package the data differently based on the requested revision of the data. In the given example the calculation "TauTotal" is only performed for revisions greater or equal 1.

```{r, echo = TRUE, eval=FALSE}
retrieveData("example", rev=1)
```

`retrieveData` will perform the calculations, create log files and package the produced files together with the log files into a compressed tgz file. The file can be found in the ouput folder of the main directory specified in the MADRaT config.


## Use own functions with MADRaT

Own functions can be made available to MADRaT just by sourcing them. They can be made visible to MADRaT by setting the option `globalenv=TRUE`. The following example shows how that can look like. 

```{r, echo = TRUE , eval=FALSE}
library(madrat)

# add global environment to madrat search path
setConfig(globalenv=TRUE)

# define simple calc-function
calcPi <- function() {
  out <- toolCountryFill(NULL,fill=pi)
  return(list(x=out,
              weight=out,
              unit="1",
              description="Just pi"))
}

# rund calcPi through wrapper function calcOutput
calcOutput("Pi")

```
In the given example `calcPi` is a calculation function which is just assigning the value pi to each country and given each country the same weight for a weighted aggregation (pi). After sourcing the function it can be used through the calc-wrapper function `calcOutput("Pi")`. The result is the aggregated data to the default region set up.

The same procedure works also for all other functions such as `downloadXYZ`, `readXYZ`, `correctXYZ`, `convertXYZ` and `fullXYZ`.


### Create MADRaT-based package

Since version 1.00 madrat allows to link packages to it and make use of its functionality. For linking madrat (in version >= 1.00) has to be added as a package dependency.

```{r, eval=FALSE}
Depends: madrat(>= 1.00)
```

In addition the following lines of code should be added as `madrat.R` to the R folder of the package. 

```{r, eval=FALSE}
### madrat.R
#' @importFrom madrat vcat

.onLoad <- function(libname, pkgname){
  madrat::setConfig(packages=c(madrat::getConfig("packages"),pkgname), .cfgchecks=FALSE, .verbose=FALSE)
}

#create an own warning function which redirects calls to vcat (package internal)
warning <- function(...) vcat(0,...)

# create a own stop function which redirects calls to stop (package internal)
stop <- function(...) vcat(-1,...)

# create an own cat function which redirects calls to cat (package internal)
cat <- function(...) vcat(1,...)
```

The `.onLoad` statement makes sure that the package is linked to madrat as soon as it is loaded. The replacements of `warning`, `stop` and `cat` are required to make use of the specific notification system in madrat, which makes for instance sure that all notes, warnings and error messages will show up in the written log files.

Besides these modifications no further changes are required and functions in the new package will be visible to the `madrat` wrapper functions.

